---
title: "Exploratory Data Analysis and Data Visualization"
author: "Preethi G C(2314578)"
date: "2024-03-31"
output:
  html_document: default
  pdf_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE)
```


```{r, echo=TRUE, include=TRUE, warning=FALSE}
library(knitr)
```


## Data Acqusition and Preliminary examination of Crime data
```{r, echo=TRUE, include=TRUE, warning=FALSE}

crime_data <- read.csv(file = "crime23.csv", TRUE , ",")
temp_data <- read.csv(file = "temp2023.csv", TRUE , ",")
```

###### Dimensions of crime data

```{r, echo=TRUE, include=TRUE, warning=FALSE}
# Display dimension of crime data
dim(crime_data)

```

##### Number of rows in crime data

```{r, echo=TRUE, include=TRUE, warning=FALSE}
# Total number of rows
nrow(crime_data)

```

##### Number of columns in crime data

```{r, echo=TRUE, include=TRUE, warning=FALSE}
# Total number of columns 
ncol(crime_data)

```


##### Columns of crime data

*Table 1: Columns of Crime data*
```{r, echo=TRUE, include=TRUE, warning=FALSE}
# Display column names of crime data
knitr::kable(data.frame(Column_names = names(crime_data)), caption = 'Column names of Crime data')
```


##### Sample of Crime dataset
*Table 2: Representation of first 6 rows of crime data*
```{r, echo=TRUE, include=TRUE, warning=FALSE}
knitr::kable(head(crime_data), caption = 'First 6 rows of Crime data')

```

##### Configuaration of crime data
```{r, echo=TRUE, include=TRUE, warning=FALSE}
configuaration_output <- capture.output(str(crime_data))
variable_names <- gsub("^[[:space:]]*([^:]+):.*$", "\\1", configuaration_output)
data_types <- gsub("^[[:space:]]*[^:]+: *([^ ]+).*", "\\1", configuaration_output)
structure_table <- data.frame('Column_Names' = variable_names , 'Data_type' = data_types)
knitr::kable(structure_table, caption = 'Configuaration of crime data')

```

##### Structure of crime data

```{r, echo=TRUE, include=TRUE, warning=FALSE}
str(crime_data)
```

```{r, echo=TRUE, include=TRUE, warning=FALSE}
mode_value <- mode(crime_data$outcome_status)
crime_data$outcome_status[is.na(crime_data$outcome_status)] <- mode_value

```

##### Missing Values of crime data

```{r, echo=TRUE, include=TRUE, warning=FALSE}
crime_data[crime_data== ""] <- NA
missing_values <- colSums(is.na(crime_data))
missing_values_df <- data.frame('Missing value census' = missing_values)
knitr::kable(missing_values_df)

```

##### Summary of Crime dataset

```{r, echo=TRUE, include=TRUE, warning=FALSE}
knitr::kable(t(summary(crime_data)), caption = 'Summary of Crime data')

```


## Data Visualization of crime data


```{r, echo=TRUE, include=TRUE, warning=FALSE}
library(kableExtra)

frequency_table <- table(crime_data$category)
frequency_df <- as.data.frame(frequency_table)
colnames(frequency_df) <- c("Category", "Frequency")
kable(frequency_df, caption = "Frequency of Crime Categories") %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE)


```


```{r, echo=TRUE, include=TRUE, warning=FALSE}
library(plotly)
library(ggplot2)

custom_colors <- c("#000033" , "#000066" , "#000099" , "#0000CC" , '#0000FF' , '#003366' , '#003399' , '#0033CC' , '#0033FF' , '#006699' , '#0066CC' , '#0066FF' , '#0099CC' , '#0099FF')

s <- ggplot(frequency_df, aes(x = Category, y = Frequency, fill = Category)) +
  geom_bar(stat = 'identity', fill = custom_colors) +
  labs(X = 'Category', Y = 'Frequency', title = 'Frequency Distribution') +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
ggplotly(s)
  
```
*Graph 1: Bar plot of Frequencies*


```{r, echo=TRUE, include=TRUE, warning=FALSE}
library(RColorBrewer)

```


```{r, echo=TRUE, include=TRUE, warning=FALSE}
library(plotly)
custom_colors <- c("#ADD8E6","blue")
pie_chart <- plot_ly(data = crime_data, labels = ~location_type, type = "pie", 
                                 marker = list(colors = custom_colors)) %>%
  layout(title = "Incident location Descriptors", showlegend = TRUE)
pie_chart

```

*Graph 2: Visualization of location types with a pie chart*


```{r, echo=TRUE, include=TRUE, warning=FALSE}
library(dplyr)
library(ggplot2)

df <- crime_data %>%
  summarise(incident_count = n())

ggplot(df, aes(x = date, y = incident_count)) +
  geom_point(color = "#0099FF", size = 3) + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Time Series of Incidents", x = "Date", y = "Number of Incidents")

```

*Graph 3: Dot plot of incidents*


```{r, echo=TRUE, include=TRUE, warning=FALSE}
library(ggplot2)
ggplot(data = crime_data, aes(x = lat)) +
  geom_histogram(fill = 'darkgrey', color = 'black', bins = 20) +
  labs(title = 'Latitude Frequency Distribution', x = 'Latitude', y = 'Frequency')

```

*Graph 4: Distribution of latitude using Histogram*


```{r, echo=TRUE, include=TRUE, warning=FALSE}
library(ggplot2)

ggplot(crime_data, aes(x = long)) +
  geom_density(color = '#0099CC', size = 1) +
  labs(title = 'Longitude illumination', x = 'Longitude', y = 'Frequency') +
  theme_minimal()

```

*Graph 5: Longitude Density plot*


```{r, echo=TRUE, include=TRUE, warning=FALSE}
library(ggplot2)
orange_palette <- colorRampPalette(c('darkorange', 'lightyellow'))

box_plot <- ggplot(data = crime_data, aes(x = category, y = long, fill = category)) +
  geom_boxplot() +  
  scale_fill_manual(values = orange_palette(14)) +
  labs(title = "Longitude trends by Crime Category", x = "Crime Category", y = "Longitude") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
box_plot

```

*Graph 6: Box plot of longitude trends*


```{r, echo=TRUE, include=TRUE, warning=FALSE}
library(ggplot2)
lavender_palette <- colorRampPalette(c('#E6E6FA', '#9370DB'))
violin_plot <- ggplot(data = crime_data, aes(x = date, y = lat, fill = date)) +
  geom_violin() +
  scale_fill_manual(values = lavender_palette(14)) +
  labs(title = 'Violin plot', x = 'Date', y = 'Latitude') +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  theme(legend.key.size = unit(0.3, "cm")) 
violin_plot

```
*Graph 7: Violin plot of latitude*


```{r, echo=TRUE, include=TRUE, warning=FALSE}
library(leaflet)

my_palette <- colorRampPalette(c("red", "blue"))
map <- leaflet(crime_data) %>%
  addTiles() %>%
  addCircleMarkers(
    radius = ~ 0.00002,
    popup = ~paste("Reason: ", crime_data$category),
    color = my_palette(14)) %>%
  addLegend("bottomleft", 
            labels = unique(crime_data$category),
            opacity = 1,
            colors = my_palette(14),
            title = "Category")

map
```

*Graph 8: Map*


## Data Analysis of Temperature Data


##### Dimenions of Temperature Data
```{r, echo=TRUE}
#Display the dimensions of temperature dataset
dim(temp_data)

```
##### Number of rows in temperature data

```{r, echo=TRUE}
nrow(temp_data)

```
##### Number of columns in temperature data

```{r, echo=TRUE}
ncol(temp_data)

```
##### Columns of temperature data

```{r, echo=TRUE, include=TRUE, warning=FALSE}
knitr::kable(data.frame(Column_names = names(temp_data)), caption = 'Column names of Temperature data')

```
##### Sample data of Temperature dataset
```{r, echo=TRUE, include=TRUE, warning=FALSE}

knitr::kable((head(temp_data)), caption = 'First 6 rows of Temperature data')

```

##### Configuaration of temperature data

```{r, echo=TRUE, include=TRUE, warning=FALSE}
configuaration_output <- capture.output(str(temp_data))
variable_names <- gsub("^[[:space:]]*([^:]+):.*$", "\\1", configuaration_output)
data_types <- gsub("^[[:space:]]*[^:]+: *([^ ]+).*", "\\1", configuaration_output)
structure_table <- data.frame('Column_Names' = variable_names , 'Data_type' = data_types)
knitr::kable(structure_table, caption = 'Configuaration of crime data')

```


##### Missing Values of Temperature data

```{r, echo=TRUE, include=TRUE, warning=FALSE}
temp_data[temp_data== ""] <- NA
missing_values <- colSums(is.na(temp_data))
missing_values_df <- data.frame('Missing value census' = missing_values)
knitr::kable(missing_values_df)

```


```{r, echo=TRUE, include=TRUE, warning=FALSE}


mean_value <- mean(temp_data$Precmm, na.rm = TRUE)
temp_data$Precmm[is.na(temp_data$Precmm)] <- mean_value

mean_value <- mean(temp_data$lowClOct, na.rm = TRUE)
temp_data$lowClOct[is.na(temp_data$lowClOct)] <- mean_value

mean_value <- mean(temp_data$SunD1h, na.rm = TRUE)
temp_data$SunD1h[is.na(temp_data$SunD1h)] <- mean_value

```


```{r, echo=TRUE, include=TRUE, warning=FALSE}
library(ggplot2)

sina_plot <- ggplot(data = temp_data, aes(x = TdAvgC, y = HrAvg)) +
  geom_violin() +
  geom_point() +
  labs(title = 'Sina plot', x = 'Average dew point', y = 'Average Relative Humidity') 

sina_plot

```

*Graph 9: Sina plot* 


```{r, echo=TRUE, include=TRUE, warning=FALSE}
scatter_plot<- ggplot(data = temp_data, aes(x = TemperatureCMin, y = TemperatureCMax, color = TemperatureCAvg)) +
  geom_point() +
  labs(title = 'Scatter plot', x = 'min temperature', y = 'max temperature') +
  scale_alpha_manual(values = c('A' = 'Blue'))
scatter_plot

```

*Graph 10: Scatter plot*


```{r, echo=TRUE, include=TRUE, warning=FALSE}
library(GGally)

data <- temp_data[, c("Precmm", "TotClOct", "lowClOct", "SunD1h")]
pair_plot <- ggpairs(data, 
                     title = 'Pair plot matrix',
                     upper = list(continous = 'cor'),
                     lower = list(continuous = 'points'),
                     diag = list(continuous = 'bar'))

pair_plot
```

*Graph 11: Pair plot matrix*



```{r, echo=TRUE, include=TRUE, warning=FALSE}
library(corrplot)

preferred_column <- c("TemperatureCAvg", "TemperatureCMax", "TemperatureCMin", "TdAvgC","HrAvg", "WindkmhInt", "WindkmhGust", "Precmm", "TotClOct", "lowClOct")
preferred_data <- temp_data[, preferred_column]
correlation_matrix <- cor(preferred_data, use = "complete.obs")
my_palette <- colorRampPalette(c('#FFD580', '#CC5500'))
corrplot(correlation_matrix, method = "color", type = "lower", tl.col = "black",
         tl.cex = 0.7, tl.srt = 45,
         title = 'Correlation matrix for temperature variables', 
         col = my_palette(10),
         addCoef.col= 'black', is.corr = TRUE, diag = TRUE, outline = TRUE, 
         order ="hclust",
         mar = c(0, 0, 2, 0),
         number.cex =  0.5)
```

*Graph 12: Correlation matrix*


```{r, echo=TRUE, include=TRUE, warning=FALSE}
library(ggplot2)

ggplot(temp_data, aes(x = Date)) +
  geom_line(aes(y = WindkmhGust), color = "#C8A2C8", size = 0.5) + 
  geom_line(aes(y = WindkmhInt), color = "#5453A6", size = 0.5) +
  labs(x = "Date", y = "Wind (km/h)", title = "Wind Over Time")

```

*Graph 13: Time series plot*

